Descriptive Complexity for Counting Complexity Classes

### 2017-03-03 LMCS/conclusions.tex
journal primera version

We proposed a framework based on Weighted Logics to develop a descriptive complexity theory for complexity classes of functions.
%In particular, we show how this framework can be used to capture fundamental counting complexity classes such as $\fp$, $\shp$ and $\fpspace$, among others. Moreover, we use it to define a hierarchy inside $\shp$, identifying counting complexity classes with good closure and approximation properties, and which admit natural complete problems. Finally, by adding recursion to the framework, we show how to capture lower counting complexity classes such as $\shl$.	
We consider the results of this paper as a first step in this direction.
%towards a descriptive complexity theory for complexity classes of functions. ç
In this sense, there are several directions for future research, some of which are mentioned here. 
$\totp$ is an important counting complexity class for which we are missing a logical characterization. In the same way, we are missing logical characterization of harder fundamental complexity class such as $\spanl$. We would also like to define a larger syntactic subclass of $\shp$ where each function admits an FPRAS; notice that $\cpm$ is an important problem admitting an FPRAS\cite{JSV04} that is not included in the classes defined in Section \ref{sec-clo}. Moreover, by following the approach proposed in
\cite{I83}, we would like to include second-order free variables in the operator for counting paths introduced in Section \ref{sec:beyond}, so to have alternative ways to capture $\fpspace$ and even $\shp$. Finally, the least fixed point operator introduced in Section \ref{sec:beyond} clearly deserves further investigation.

>>>>

We proposed a framework based on Weighted Logics to develop a descriptive complexity theory for complexity classes of functions.
%In particular, we show how this framework can be used to capture fundamental counting complexity classes such as $\fp$, $\shp$ and $\fpspace$, among others. Moreover, we use it to define a hierarchy inside $\shp$, identifying counting complexity classes with good closure and approximation properties, and which admit natural complete problems. Finally, by adding recursion to the framework, we show how to capture lower counting complexity classes such as $\shl$.	
We consider the results of this paper as a first step in this direction.
%towards a descriptive complexity theory for complexity classes of functions. ç
In this sense, there are several directions for future research, some of which are mentioned here. 
$\totp$ is an interesting counting complexity class as it naturally defines a class of functions in $\shp$ with easy decision counterparts. However, we do not have a logical characterization of this class.
%for which we are missing a logical characterization. 
In the same direction, we are missing logical characterizations of other fundamental complexity classes such as $\spanl$. We would also like to define a larger syntactic subclass of $\shp$ where each function admits an FPRAS; notice that $\cpm$ is an important problem admitting an FPRAS\cite{JSV04} that is not included in the classes defined in Section \ref{sec-clo}. Moreover, by following the approach proposed in
\cite{I83}, we would like to include second-order free variables in the operator for counting paths introduced in Section \ref{sec:beyond}, so to have alternative ways to capture $\fpspace$ and even $\shp$. Finally, the least fixed point operator introduced in Section \ref{sec:beyond} clearly deserves further investigation.

### 2017-03-03 LMCS/counting.tex
journal primera version

Fix a signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered (finite) $\R$-structure with a domain $A = \{a_1, \ldots, a_n\}$.
%$ containing $n$ elements. 
Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the string
%following binary string~\cite{L04}:
%\begin{eqnarray*}
	%\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
%\end{eqnarray*}
$0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A)$~\cite{L04}. 
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a counting complexity class. From now on, we assume that every counting complexity class $\CC$ consists of functions $f : \{0,1\}^* \to \N$ that are invariant under isomorphism of structures. More precisely, given a signature $\R$, a function $f : \{0,1\}^* \to \N$ is invariant under $\R$-isomorphisms if for every pair $\A, \mathfrak{B} \in \ostr[\R]$ such that $\A$ and 
$\mathfrak{B}$ are isomorphic, it holds that $f(\enc(\A)) = f(\enc(\mathfrak{B}))$. It is worth noticing that this is a usual assumption in descriptive complexity~\cite{L04}.
%over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a counting complexity class. Then {\em  $\FF$ captures $\CC$ over ordered $\R$-structures} if the  following conditions hold:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$ that is invariant under $\R$-isomorphisms, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Moreover, {\em $\FF$ captures $\CC$ over ordered structures} if $\FF$ captures~$\CC$ over ordered $\R$-structures for every signature~$\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
In Definition~\ref{def:cap}, function $f \in \CC$ and formula $\alpha \in \FF$ must coincide in all the strings that encode ordered $\R$-structures. Notice that this restriction is natural as we want to capture %Since we want to capture 
$\CC$ over a fix set of structures (e.g. graphs, matrices).
%, it is natural to just consider strings that encodes $\R$-structures. 
Moreover, this restriction is fairly standard in descriptive complexity \cite{immerman1999descriptive,L04}, and it has also been used in the previous work on capturing complexity classes of functions \cite{SalujaST95,ComptonG96}.
%all notions for capturing complexity classes restrict $f \in \CC$ similarly.

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{F75} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} follows by the fact that $\shp$ is closed under first- and second-order sum.
%Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
%\begin{corollary} \label{prop:capture-shP-LFP}
%	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
%\end{corollary}

By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Thus, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and product quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

At this point it is natural to ask whether one can extend the previous idea to capture $\fpspace$~\cite{Ladner89}, the class of functions computable in polynomial space. 
%Another natural function complexity class is $\fpspace$~\cite{Ladner89}, the functional version of $\pspace$. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of a polynomial-space TM.
% $\pspace$-machine. 
Moreover, 
%On the other hand, 
one also needs more powerful quantitative quantifiers as functions like $2^{2^n}$ can be computed in polynomial space,
%such a TM can generatean $\fpspace$ machine can generate outputs of exponentially many bits. 
so second-order sum is not enough for the quantitative layer of a logic for $\fpspace$.
%For this reason, the second-order sum is not enough for the quantitative layer of a logic for $\fpspace$, that is, one needs to consider second-order products to produce big numbers. 
In fact, by considering second-order product we obtain the fragment $\qso(\pfp)$ that 
%naturally 
captures $\fpspace$. 
\begin{theorem} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}
The proof of the previous theorem follows the same line as for the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of a polynomial-space TM can be reconstructed by using $\pfp$ and quantitative quantifiers.

From the proof of the previous theorem a natural question follows: what happens if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also introduced the class $\nfpspace$ of all functions computed by polynomial-space TMs 
%$\pspace$-machines 
with output length bounded by a polynomial.
% output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{cor:capture-fpspace-poly}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}

The results of this section validate $\qso$ as an appropriate logical framework for extending the theory of descriptive complexity to counting complexity classes. In the following sections, we provide more arguments for this claim, by considering some fragments of $\eqso$ and, moreover, by showing how to go beyond $\eqso$ to capture other 
%counting complexity 
classes.

>>>>

Fix a signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered (finite) $\R$-structure with a domain $A = \{a_1, \ldots, a_n\}$.
%$ containing $n$ elements. 
Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the string
%following binary string~\cite{L04}:
%\begin{eqnarray*}
	%\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
%\end{eqnarray*}
$0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A)$~\cite{L04}. 
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a counting complexity class. From now on, we assume that every counting complexity class $\CC$ consists of functions $f : \{0,1\}^* \to \N$ that are invariant under isomorphism of structures. More precisely, given a signature $\R$, a function $f : \{0,1\}^* \to \N$ is invariant under $\R$-isomorphisms if for every pair $\A, \mathfrak{B} \in \ostr[\R]$ such that $\A$ and 
$\mathfrak{B}$ are isomorphic, it holds that $f(\enc(\A)) = f(\enc(\mathfrak{B}))$. It is worth noticing that this is a usual assumption in descriptive complexity~\cite{L04}.
%over $\R$-structures.
\begin{defi} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a counting complexity class. Then {\em  $\FF$ captures $\CC$ over ordered $\R$-structures} if the  following conditions hold:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$ that is invariant under $\R$-isomorphisms, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Moreover, {\em $\FF$ captures $\CC$ over ordered structures} if $\FF$ captures~$\CC$ over ordered $\R$-structures for every signature~$\R$.
\end{defi}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
In Definition~\ref{def:cap}, function $f \in \CC$ and formula $\alpha \in \FF$ must coincide in all the strings that encode ordered $\R$-structures. Notice that this restriction is natural as we want to capture %Since we want to capture 
$\CC$ over a fix set of structures (e.g. graphs, matrices).
%, it is natural to just consider strings that encodes $\R$-structures. 
Moreover, this restriction is fairly standard in descriptive complexity \cite{immerman1999descriptive,L04}, and it has also been used in the previous work on capturing complexity classes of functions \cite{SalujaST95,ComptonG96}.
%all notions for capturing complexity classes restrict $f \in \CC$ similarly.

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{F75} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected.

For a given signature $\R$, we define $\ostr[\R]^*$ as
\begin{multline*}
\ostr[\R]^* = \{(\A,v,V) \mid \A\in\ostr[\R]\text{,} \\ \text{$v$ ($V$) is a first-order (second-order) assignment for $\A$}  \}.
\end{multline*}

\begin{prop} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{prop}
\proof
\input{proof-capture-shP} 
\qed
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} follows by the fact that $\shp$ is closed under first- and second-order sum.

By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{prop} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{prop}
\proof
\input{proof-capture-spanP}
\qed
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
Thus, it is not  clear how to capture $\fp$ 
by following the approach proposed in~\cite{SalujaST95}. 
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.

From now on, we will use the symbol $<$ also to denote the lexicographic order over same-sized tuples. If $\bar{x} = (x_1,\ldots,x_m)$ and $\bar{y} = (y_1,\ldots,y_m)$ are tuples of first-order variables, we denote $\bar{x} < \bar{y}$ for the formula $\bigvee_{i = 1}^m[\bigwedge_{j = 1}^{i-1}x_j = y_j \wedge x_i < y_i]$. Similarly, we use $=$ to denote equality between tuples, as $\bar{x} = \bar{y}$ denotes $\bigwedge_{i = 1}^m(x_i = y_i)$, and also $\bar{x}\leq\bar{y}$ denotes $\bar{x} < \bar{y} \vee \bar{x} = \bar{y}$. Furthermore, If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$.

The {\em conditional count} symbol $(\varphi \mapsto \alpha)$ is defined as $(\neg\varphi + (\varphi\cdot\alpha))$ for given $\so$ formula $\varphi$ and $\qso$ formula $\alpha$. Note that for each $(\A,v,V) \in \ostr[\R]^*$, 
\[
\sem{(\varphi \mapsto \alpha)}(\A,v,V) = 
\begin{cases}
\sem{\alpha}(\A,v,V) &\text{if } (\A,v,V)\models\varphi,\\
0 &\text{otherwise}.
\end{cases}
\]
\begin{thm} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{thm}
\proof
\input{proof-capture-fp}
\qed
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
It is important to notice that the alternation between sum and product quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.

At this point it is natural to ask whether one can extend the previous idea to capture $\fpspace$~\cite{Ladner89}, the class of functions computable in polynomial space. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of a polynomial-space TM.
Moreover, 
one also needs more powerful quantitative quantifiers as functions like $2^{2^n}$ can be computed in polynomial space,
so second-order sum is not enough for the quantitative layer of a logic for $\fpspace$.
In fact, by considering second-order product we obtain the fragment $\qso(\pfp)$ that captures $\fpspace$. 
\begin{thm} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{thm}
\proof
\input{proof-capture-fpspace}
\qed
The proof of the previous theorem follows the same line as for the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of a polynomial-space TM can be reconstructed by using $\pfp$ and quantitative quantifiers.

From the proof of the previous theorem a natural question follows: what happens if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also introduced the class $\nfpspace$ of all functions computed by polynomial-space TMs 
with output length bounded by a polynomial.
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{cor} \label{cor:capture-fpspace-poly}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{cor}
\proof
\input{proof-capture-fpspace-poly}
\qed

The results of this section validate $\qso$ as an appropriate logical framework for extending the theory of descriptive complexity to counting complexity classes. In the following sections, we provide more arguments for this claim, by considering some fragments of $\eqso$ and, moreover, by showing how to go beyond $\eqso$ to capture other classes.

### 2017-03-03 LMCS/extentions.tex
journal primera version

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
%such that for some $\np$ machine, $f(x)$ is the number of accepting runs minus the number of rejecting runs.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
%it can compute negative values by definition. 
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to %describe functions
capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{example}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{example}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{corollary} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{corollary}
%The previous result is straightforward to prove from the characterization of $\shp$ and the definition of $\gp$. 
%However, 
%We 
%believe 
%think
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

@@

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
%operators of first- and second-order. 
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
%the formulae $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. 
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{exa}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{exa}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{cor} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{cor}
This is an interesting result that shows how robust is $\qso$ when capturing different complexity classes.

@@

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{align*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{align*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum. 
\begin{exa}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{exa}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{thm} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{thm}
\proof
\input{proof-capture-optp}
\qed
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.
